---
title: "GenBank reader benchmark"
output:
  github_document:
    html_preview: false
    pandoc_args: ["--wrap=preserve", "--columns=400"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

This report measures `read_genbank(...)` and `genbank_to_fasta(...)` on a real
annotation: the NCBI RefSeq record for *Escherichia coli* K-12 MG1655
(`NC_000913.3`), a 4,641,652 bp circular genome yielding 9,306 feature rows
across 8 feature kinds. Fetching the record is staging work and is outside every
timed interval.

Two builds are timed on the same host in one render: the baseline is the reader
as reviewed at f855f881, which emitted rows while streaming, and the candidate is
the record-level parser that replaced it. The baseline and candidate read the
same staged input under the same CPU affinity, each through its own in-memory
DuckDB connection, baseline first.

There is no earlier checked-in DuckHTS GenBank benchmark besides the previous
render of this report. The nearest reader report is
[`benchmark_gffbase_conformance.md`](benchmark_gffbase_conformance.md), which
measures `read_gff(...)`. It is the closest analogue because `read_genbank`
emits the exact column shape of `read_gff`, but the two are not numerically
comparable: that workload reads tabix-style tabular rows, while this parses a
GenBank flat file line by line through hFILE.

`read_genbank` declares no parallel scan and streams one record at a time, so
the one- and four-thread conditions exist to show whether threading adds
overhead rather than to show it scales. Memory tracks the feature table of
the largest record rather than the projection; that limit is visible in the input
denominators below but is not itself measured here.

## Reproduction

The input is the duckhtsbench-registered NCBI RefSeq assembly archive
`GCF_000005845.2_ASM584v2_genomic.gbff.gz` (workload `genbank-reader`), pinned
by the MD5 NCBI publishes, its SHA-256 and byte size, with the uncompressed
`.gbff` the report reads registered as a derived artifact with its own
SHA-256. NCBI re-annotates RefSeq assemblies in place, so a later archive that
no longer matches the registry is refused at staging rather than silently
measured. Stage it once, with network access only in this step:

``` sh
Rscript -e 'duckhtsbench::duckhts_bench_stage_genbank()'
```

Build the baseline revision separately, for example in a worktree, and point
`DUCKHTS_GENBANK_BASELINE_EXTENSION` and `DUCKHTS_GENBANK_BASELINE_REVISION` at
its binary and commit. `DUCKHTS_EXTENSION` selects the candidate (default:
`build/release/duckhts.duckdb_extension`). Then render from a tracked-clean
revision; rendering resolves the staged path through the registry and does not
download:

``` sh
git worktree add ../duckhts-baseline f855f881
(cd ../duckhts-baseline && make configure && make release)
DUCKHTS_GENBANK_BASELINE_EXTENSION=../duckhts-baseline/build/release/duckhts.duckdb_extension \
DUCKHTS_GENBANK_BASELINE_REVISION=f855f881 \
  taskset -c 8-11 Rscript -e 'rmarkdown::render("benchmarks/benchmark_genbank_reader.Rmd")'
```

The benchmark pins the one-thread condition to the highest logical CPU in the
affinity mask of the rendering process and the four-thread condition to the highest
four, so wrap the render in `taskset -c` to choose them; the table records the
affinity each condition ran under. Input staging, extension loading,
connection setup, warm-up, result verification, and removal of the previous
FASTA output are excluded from timing.

```{r benchmark, echo=FALSE, message=FALSE}
suppressPackageStartupMessages({
  library(DBI)
  library(duckdb)
})

sql_string <- function(x) {
  sprintf("'%s'", gsub("'", "''", x, fixed = TRUE))
}

command_output <- function(command, args = character()) {
  paste(system2(command, args, stdout = TRUE, stderr = TRUE), collapse = " ")
}

input_path <- duckhtsbench::duckhts_bench_stage_genbank(fetch = FALSE)[["gbff"]]
input_path <- normalizePath(input_path, mustWork = TRUE)

baseline_extension <- Sys.getenv("DUCKHTS_GENBANK_BASELINE_EXTENSION")
baseline_revision <- Sys.getenv("DUCKHTS_GENBANK_BASELINE_REVISION")
candidate_extension <- Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension")
stopifnot(
  nzchar(baseline_revision),
  file.exists(baseline_extension),
  file.exists(candidate_extension)
)
builds <- c(
  baseline = normalizePath(baseline_extension),
  candidate = normalizePath(candidate_extension)
)

candidate_revision <- command_output("git", c("rev-parse", "--short=12", "HEAD"))
candidate_src_tree <- command_output("git", c("rev-parse", "HEAD:src"))
tracked_dirty <- length(system2(
  "git", c("status", "--porcelain", "--untracked-files=no"), stdout = TRUE
)) > 0L
if (tracked_dirty) {
  stop("Render the benchmark from a tracked-clean source revision")
}

open_build <- function(extension) {
  con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
  invisible(dbExecute(con, paste("LOAD", dbQuoteString(con, extension))))
  con
}

# One row back to R, so the timing covers parsing and decoding rather than
# transport. The fingerprint is order independent, so a reordered scan that
# still produced the right bag of rows would pass, and a dropped or duplicated
# feature would not.
aggregate_sql <- function(attributes_map = FALSE) {
  # The scan only fills PROJECTED columns, so selecting attributes_map is what makes
  # the MAP get built at all. Aggregating its cardinality keeps one row coming back
  # while forcing every entry to be materialized.
  map_term <- if (attributes_map) {
    "sum(cardinality(attributes_map))::DOUBLE AS map_entries"
  } else {
    "0::DOUBLE AS map_entries"
  }
  sprintf(
    paste0(
      "SELECT count(*)::UBIGINT AS rows, ",
      "hex(bit_xor(hash(seqname, feature, start, \"end\", strand, frame, attributes))) ",
      "AS fingerprint, sum(\"end\" - start + 1)::DOUBLE AS span_sum, %s ",
      "FROM read_genbank(%s, attributes_map := %s)"
    ),
    map_term, sql_string(input_path), if (attributes_map) "TRUE" else "FALSE"
  )
}

# Each build is verified against its own answer: the baseline is the reviewed
# reader and emits different phases and attribute text by design, so the two
# builds share the row count and the input but not the fingerprint.
build_expectations <- function(con) {
  plain <- dbGetQuery(con, aggregate_sql(FALSE))
  map <- dbGetQuery(con, aggregate_sql(TRUE))
  stopifnot(as.double(plain$rows[[1L]]) > 9000)
  # The MAP projection must not change which features are emitted.
  stopifnot(identical(plain$fingerprint[[1L]], map$fingerprint[[1L]]))
  # The MAP workload must actually have built entries, or it is measuring a pruned column.
  stopifnot(map$map_entries[[1L]] > plain$rows[[1L]])
  list(plain = plain, map = map)
}

same_answer <- function(observed, expected) {
  identical(as.character(observed$rows[[1L]]), as.character(expected$rows[[1L]])) &&
    identical(observed$fingerprint[[1L]], expected$fingerprint[[1L]]) &&
    isTRUE(all.equal(
      observed$span_sum[[1L]], expected$span_sum[[1L]],
      tolerance = 1e-12
    )) &&
    isTRUE(all.equal(
      observed$map_entries[[1L]], expected$map_entries[[1L]],
      tolerance = 1e-12
    ))
}

set_affinity <- function(cpus) {
  output <- system2(
    "taskset", c("-pc", cpus, Sys.getpid()),
    stdout = TRUE, stderr = TRUE
  )
  if (!is.null(attr(output, "status"))) stop(paste(output, collapse = " "))
  invisible(output)
}

# The CPUs this render may use, as taskset reports them, so an outer
# `taskset -c` chooses the measurement CPUs without editing this file.
affinity_cpus <- function() {
  line <- system2("taskset", c("-pc", Sys.getpid()), stdout = TRUE)
  spec <- sub("^.*: *", "", line[[length(line)]])
  unlist(lapply(strsplit(spec, ",", fixed = TRUE)[[1L]], function(part) {
    bounds <- as.integer(strsplit(part, "-", fixed = TRUE)[[1L]])
    if (length(bounds) == 1L) bounds else seq.int(bounds[[1L]], bounds[[2L]])
  }))
}
host_cpus <- affinity_cpus()
if (length(host_cpus) < 4L) stop("the render needs at least four logical CPUs in its affinity mask")
cpu_one <- as.character(host_cpus[[length(host_cpus)]])
cpu_four <- paste(host_cpus[(length(host_cpus) - 3L):length(host_cpus)], collapse = ",")

measure_aggregate <- function(con, workload, attributes_map, expected, threads, cpus,
                              repetitions = 9L) {
  set_affinity(cpus)
  invisible(dbExecute(con, sprintf("SET threads = %d", threads)))
  sql <- aggregate_sql(attributes_map)
  for (i in seq_len(2L)) stopifnot(same_answer(dbGetQuery(con, sql), expected))
  seconds <- numeric(repetitions)
  for (i in seq_len(repetitions)) {
    started <- proc.time()[["elapsed"]]
    observed <- dbGetQuery(con, sql)
    seconds[[i]] <- proc.time()[["elapsed"]] - started
    stopifnot(same_answer(observed, expected))
  }
  data.frame(
    workload = workload,
    threads = threads,
    affinity = cpus,
    repetitions = repetitions,
    output_rows = as.double(expected$rows[[1L]]),
    output_bytes = NA_real_,
    min_seconds = min(seconds),
    median_seconds = median(seconds),
    max_seconds = max(seconds),
    median_rows_per_second = as.double(expected$rows[[1L]]) / median(seconds),
    fingerprint = expected$fingerprint[[1L]],
    span_sum = expected$span_sum[[1L]],
    stringsAsFactors = FALSE
  )
}

# The write path reads ORIGIN rather than FEATURES, so it exercises the other
# half of the parser. Each pass is verified by reading the FASTA back.
measure_fasta <- function(con, threads, cpus, repetitions = 5L) {
  set_affinity(cpus)
  invisible(dbExecute(con, sprintf("SET threads = %d", threads)))
  output_path <- tempfile(sprintf("genbank-t%d-", threads), fileext = ".fa")
  convert_sql <- sprintf(
    "SELECT records_written FROM genbank_to_fasta(%s, output_path := %s, overwrite := TRUE)",
    sql_string(input_path), sql_string(output_path)
  )
  verify_sql <- sprintf(
    "SELECT count(*)::UBIGINT AS rows, sum(length(SEQUENCE))::DOUBLE AS bases FROM read_fasta(%s)",
    sql_string(output_path)
  )
  seconds <- numeric(repetitions)
  output_bytes <- numeric(repetitions)
  records <- NA_real_
  for (i in seq_len(repetitions)) {
    unlink(output_path)
    started <- proc.time()[["elapsed"]]
    written <- dbGetQuery(con, convert_sql)
    seconds[[i]] <- proc.time()[["elapsed"]] - started
    output_bytes[[i]] <- file.info(output_path)$size
    records <- as.double(written$records_written[[1L]])
    verified <- dbGetQuery(con, verify_sql)
    stopifnot(as.double(verified$rows[[1L]]) == records)
    stopifnot(as.double(verified$bases[[1L]]) == 4641652)
  }
  unlink(output_path)
  data.frame(
    workload = "ORIGIN to FASTA",
    threads = threads,
    affinity = cpus,
    repetitions = repetitions,
    output_rows = records,
    output_bytes = median(output_bytes),
    min_seconds = min(seconds),
    median_seconds = median(seconds),
    max_seconds = max(seconds),
    median_rows_per_second = 4641652 / median(seconds),
    fingerprint = NA_character_,
    span_sum = NA_real_,
    stringsAsFactors = FALSE
  )
}

results <- list()
expectations <- list()
versions <- list()
for (build in names(builds)) {
  con <- open_build(builds[[build]])
  expected <- build_expectations(con)
  expectations[[build]] <- expected$plain
  versions[[build]] <- c(
    duckdb = dbGetQuery(con, "SELECT version() AS v")$v[[1L]],
    htslib = dbGetQuery(con, "SELECT duckhts_htslib_version() AS v")$v[[1L]],
    kinds = as.character(dbGetQuery(
      con,
      sprintf("SELECT count(DISTINCT feature) AS k FROM read_genbank(%s)", sql_string(input_path))
    )$k[[1L]])
  )
  rows <- rbind(
    measure_aggregate(con, "features + aggregate", FALSE, expected$plain, 1L, cpu_one),
    measure_aggregate(con, "features + aggregate", FALSE, expected$plain, 4L, cpu_four),
    measure_aggregate(con, "features + attribute MAP", TRUE, expected$map, 1L, cpu_one),
    measure_aggregate(con, "features + attribute MAP", TRUE, expected$map, 4L, cpu_four),
    measure_fasta(con, 1L, cpu_one),
    measure_fasta(con, 4L, cpu_four)
  )
  results[[build]] <- cbind(build = build, rows, stringsAsFactors = FALSE)
  invisible(dbDisconnect(con, shutdown = TRUE))
}
results <- do.call(rbind, results)
rownames(results) <- NULL

# Both builds must emit the same number of feature rows over the same input, or
# the timings are not for the same work.
stopifnot(identical(
  as.character(expectations$baseline$rows[[1L]]),
  as.character(expectations$candidate$rows[[1L]])
))

results$baseline_revision <- baseline_revision
results$candidate_revision <- candidate_revision
results$run_date <- as.character(Sys.Date())
results$input_path <- input_path
results$input_bytes <- file.info(input_path)$size
results$input_span_bases <- 4641652
results$duckdb_version <- versions$candidate[["duckdb"]]
results$htslib_version <- versions$candidate[["htslib"]]

dir.create("benchmarks/data", showWarnings = FALSE, recursive = TRUE)
write.csv(
  results,
  "benchmarks/data/genbank_reader.csv",
  row.names = FALSE,
  na = ""
)

cpu_model <- sub(
  "^model name[[:space:]]*:[[:space:]]*", "",
  grep("^model name", readLines("/proc/cpuinfo"), value = TRUE)[[1L]]
)
metadata <- data.frame(
  property = c(
    "baseline revision", "candidate revision", "candidate src tree",
    "baseline binary MD5", "candidate binary MD5", "run date",
    "input source", "input accession", "GenBank bytes", "genome bases",
    "emitted feature rows", "distinct feature kinds",
    "baseline feature fingerprint", "candidate feature fingerprint",
    "DuckDB version", "htslib version", "CPU"
  ),
  value = c(
    baseline_revision,
    candidate_revision,
    candidate_src_tree,
    unname(tools::md5sum(builds[["baseline"]])),
    unname(tools::md5sum(builds[["candidate"]])),
    as.character(Sys.Date()),
    "duckhtsbench registry genbank-reader: NCBI RefSeq GCF_000005845.2_ASM584v2 genomic.gbff.gz, gunzipped",
    "NC_000913.3",
    format(file.info(input_path)$size, big.mark = ",", scientific = FALSE),
    format(4641652, big.mark = ",", scientific = FALSE),
    format(as.double(expectations$candidate$rows[[1L]]), big.mark = ",", scientific = FALSE),
    versions$candidate[["kinds"]],
    expectations$baseline$fingerprint[[1L]],
    expectations$candidate$fingerprint[[1L]],
    versions$candidate[["duckdb"]],
    versions$candidate[["htslib"]],
    cpu_model
  ),
  stringsAsFactors = FALSE
)

knitr::kable(metadata, col.names = c("Property", "Recorded value"))
```

## Results

The aggregate workloads parse the FEATURES table and return one row to R, so
the timing covers parsing and decoding rather than transport. The FASTA
workload reads ORIGIN instead and writes every base as a real file. Every timed
pass is verified against its own build: the aggregates by exact row count and an
order-independent full-row fingerprint, with the summed span checked to a
relative tolerance of `1e-12`, and the FASTA by reading it back and asserting
both the record count and 4,641,652 bases.

The two builds agree on the number of feature rows and on the bases written,
but not on the fingerprint or the FASTA byte count. The baseline is the reader
as reviewed, so its rows carry the phases, `Parent` links and attribute text
the review found wrong, and its defline keeps the period that ends the
DEFINITION. The comparison is therefore of the time each build takes to read
the same input and emit the same number of rows, not of identical output.

```{r results-table, echo=FALSE}
display <- results[, c(
  "build", "workload", "threads", "affinity", "repetitions", "output_rows",
  "output_bytes", "min_seconds", "median_seconds", "max_seconds",
  "median_rows_per_second"
)]
display$output_rows <- format(
  display$output_rows, big.mark = ",", scientific = FALSE
)
display$output_bytes <- ifelse(
  is.na(display$output_bytes), "–",
  format(display$output_bytes, big.mark = ",", scientific = FALSE)
)
for (column in c("min_seconds", "median_seconds", "max_seconds")) {
  display[[column]] <- sprintf("%.3f", display[[column]])
}
display$median_rows_per_second <- format(
  round(display$median_rows_per_second), big.mark = ",", scientific = FALSE
)
names(display) <- c(
  "Build", "Workload", "Threads", "CPU affinity", "Runs", "Output rows",
  "FASTA bytes", "Minimum seconds", "Median seconds", "Maximum seconds",
  "Median rows/s"
)
knitr::kable(display, align = c("l", "l", rep("r", 9)))
```

```{r comparison-table, echo=FALSE}
comparison <- merge(
  results[results$build == "baseline", c("workload", "threads", "median_seconds")],
  results[results$build == "candidate", c("workload", "threads", "median_seconds")],
  by = c("workload", "threads"), suffixes = c("_baseline", "_candidate")
)
comparison <- comparison[order(match(comparison$workload, unique(results$workload)), comparison$threads), ]
comparison$percent_change <- 100 * (comparison$median_seconds_candidate - comparison$median_seconds_baseline) /
  comparison$median_seconds_baseline
comparison$median_seconds_baseline <- sprintf("%.3f", comparison$median_seconds_baseline)
comparison$median_seconds_candidate <- sprintf("%.3f", comparison$median_seconds_candidate)
comparison$percent_change <- sprintf("%.1f", comparison$percent_change)
names(comparison) <- c(
  "Workload", "Threads", "Median seconds baseline", "Median seconds candidate", "Percent change"
)
rownames(comparison) <- NULL
knitr::kable(comparison, align = c("l", rep("r", 4)))
```

The features and ORIGIN rows count different things: the aggregate workloads
emit one row per feature, while the FASTA rate is expressed in bases written,
since the converter produces a single record for this genome. Adding
`attributes_map := TRUE` builds a `MAP<VARCHAR,VARCHAR>` per feature on top of
the raw attribute string and is the more expensive projection, which is why
both are recorded.

The measurement does not include the NCBI fetch, extension loading, or any
downstream join against the emitted intervals. It also does not bound memory:
the candidate streams one record at a time, so memory follows the largest
record, and this genome is a single record. A multi-record annotation is a
different question than this report answers. Small timings vary with
scheduling; the percent changes are for this host and input only.
