---
title: "Somalier site extraction: 17,000-site VCF/BCF and BAM/CRAM materialization"
output:
  github_document:
    html_preview: false
---

This report measures complete count-relation materialization for both Somalier
extraction surfaces. `duckhts_somalier_vcf_counts()` reads two encodings of the
same registered GIAB HG002 input. Its untimed panel contains 17,000 distinct
canonical biallelic SNVs selected deterministically from those records.
`duckhts_somalier_bam_counts()` reads a separately registered synthetic
17,000-site alignment workload in BAM and CRAM form with worker counts 1, 2,
and 4. Neither panel is the published Somalier population panel, and this
report does not establish biological panel quality.

The VCF.gz and BCF artifacts contain the same 319,349 chr1 records, one sample,
and all original fields including `FORMAT/AD`. Each timed query scans one
encoding with zero HTSlib decompression workers and materializes all 17,000
output rows in DuckDB. Panel construction, connection setup, extension loading,
warm-up, result verification, and VCF-versus-BCF comparison are outside the
timer. Peak resident memory is the Linux process high-water mark after the
timed materialization, so it includes that process's R, DuckDB, extension,
panel, and warm-up state rather than pretending to isolate one operator.

The synthetic alignment workload has 22 contigs, 17,000 panel sites, and
17,000 coordinate-sorted 101-base source reads: exactly one read covers each
site. Even site indices observe panel allele A and odd indices observe panel
allele B. The BAM and CRAM contain the same records and use the same generated
reference and explicit indexes. This controlled workload measures sparse
panel extraction mechanics; it is not a whole-genome I/O, coverage, or
biological accuracy workload.

## Reproduction

Stage the registered inputs explicitly, build the candidate extension, and
render from the repository root. Rendering performs no download:

```sh
DUCKHTSBENCH_REGISTRY=r/duckhtsbench/inst/benchmark_registry.tsv \
  Rscript -e 'library(duckhtsbench); duckhts_bench_stage_genotype_phase_set()'
DUCKHTSBENCH_REGISTRY=r/duckhtsbench/inst/benchmark_registry.tsv \
  Rscript -e 'library(duckhtsbench); duckhtsbench:::duckhts_bench_stage_somalier_site_extraction()'
make release -j2
DUCKHTSBENCH_REGISTRY=r/duckhtsbench/inst/benchmark_registry.tsv \
DUCKHTS_EXTENSION=build/release/duckhts.duckdb_extension \
  Rscript -e 'rmarkdown::render("benchmarks/benchmark_somalier_site_extraction.Rmd")'
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r measurement}
required_packages <- c("DBI", "duckdb", "callr", "digest", "duckhtsbench")
missing_packages <- required_packages[!vapply(required_packages, requireNamespace,
                                               logical(1L), quietly = TRUE)]
if (length(missing_packages)) {
  stop("missing benchmark package(s): ", paste(missing_packages, collapse = ", "))
}

if (!nzchar(Sys.getenv("DUCKHTSBENCH_REGISTRY"))) {
  Sys.setenv(DUCKHTSBENCH_REGISTRY = normalizePath(
    "r/duckhtsbench/inst/benchmark_registry.tsv", mustWork = TRUE))
}
extension <- normalizePath(Sys.getenv(
  "DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension"), mustWork = TRUE)
plan <- duckhtsbench::duckhts_bench_stage_plan("genotype-phase-set")
ids <- c(VCF = "geno_giab_phased_chr1_vcfgz",
         BCF = "geno_giab_phased_chr1_bcf")
if (!identical(plan$id[3:4], unname(ids)) ||
    any(!grepl("benchmark_somalier_site_extraction.Rmd",
               plan$consumer, fixed = TRUE))) {
  stop("genotype phase-set registry does not declare this benchmark consumer")
}
inputs <- vapply(ids, duckhtsbench::duckhts_bench_artifact_path, character(1L))
receipts <- paste0(inputs, ".provenance.tsv")
if (!all(file.exists(c(inputs, receipts)))) {
  stop("stage the genotype-phase-set workload before rendering")
}
read_receipt <- function(path) {
  fields <- utils::read.delim(path, colClasses = "character", check.names = FALSE,
                              stringsAsFactors = FALSE)
  if (!identical(names(fields), c("field", "value")) || anyDuplicated(fields$field)) {
    stop("malformed benchmark receipt: ", path)
  }
  stats::setNames(fields$value, fields$field)
}
receipt_fields <- lapply(receipts, read_receipt)
names(receipt_fields) <- names(inputs)
for (format in names(inputs)) {
  receipt <- receipt_fields[[format]]
  if (!identical(receipt[["artifact_id"]], ids[[format]]) ||
      !identical(receipt[["records"]], "319349") ||
      !identical(receipt[["samples"]], "1") ||
      !identical(receipt[["observed_sha256"]],
                 digest::digest(file = inputs[[format]], algo = "sha256"))) {
    stop(format, " artifact or receipt does not match the registered workload")
  }
}
if (!identical(receipt_fields$VCF[c("records", "samples", "calls", "allele_slots")],
               receipt_fields$BCF[c("records", "samples", "calls", "allele_slots")])) {
  stop("registered VCF and BCF denominators differ")
}

alignment_plan <- duckhtsbench::duckhts_bench_stage_plan("somalier-site-extraction")
alignment_ids <- c(
  panel = "somalier_site_panel",
  reference = "somalier_site_reference",
  reference_fai = "somalier_site_reference_fai",
  BAM = "somalier_site_bam",
  bam_bai = "somalier_site_bam_bai",
  CRAM = "somalier_site_cram",
  cram_crai = "somalier_site_cram_crai"
)
if (!identical(alignment_plan$id, unname(alignment_ids)) ||
    any(!grepl("benchmark_somalier_site_extraction.Rmd",
               alignment_plan$consumer, fixed = TRUE))) {
  stop("Somalier site-extraction registry does not declare this benchmark consumer")
}
alignment_inputs <- vapply(
  alignment_ids, duckhtsbench::duckhts_bench_artifact_path, character(1L)
)
alignment_receipts <- paste0(alignment_inputs, ".provenance.tsv")
if (!all(file.exists(c(alignment_inputs, alignment_receipts)))) {
  stop("stage the Somalier site-extraction workload before rendering")
}
alignment_receipt_fields <- lapply(alignment_receipts, read_receipt)
names(alignment_receipt_fields) <- names(alignment_inputs)
for (name in names(alignment_inputs)) {
  receipt <- alignment_receipt_fields[[name]]
  if (!identical(receipt[["artifact_id"]], alignment_ids[[name]]) ||
      !identical(receipt[["panel_sites"]], "17000") ||
      !identical(receipt[["source_reads"]], "17000") ||
      !identical(receipt[["artifact_bytes"]],
                 as.character(file.info(alignment_inputs[[name]])$size))) {
    stop(name, " alignment artifact or receipt does not match the registered workload")
  }
}

git_output <- function(args) {
  value <- system2("git", args, stdout = TRUE, stderr = TRUE)
  status <- attr(value, "status")
  if (!is.null(status) && status != 0L) stop("git command failed")
  value
}
revision <- git_output(c("rev-parse", "HEAD"))[[1L]]
source_changes <- git_output(c("status", "--short", "--", "src"))
source_state <- if (length(source_changes)) {
  "dirty; benchmark must be rerendered after committing"
} else {
  "clean at recorded revision"
}
cpu <- readLines("/proc/cpuinfo", warn = FALSE)
cpu <- sub("^model name[[:space:]]*:[[:space:]]*", "",
           cpu[grepl("^model name", cpu)][[1L]])
identity <- data.frame(
  property = c("candidate checkout revision", "candidate source state",
               "candidate extension", "R",
               "DuckDB R package", "HTSlib", "CPU", "DuckDB threads",
               "HTSlib decompression workers", "timing repetitions"),
  value = c(revision, source_state, extension, R.version.string,
            as.character(utils::packageVersion("duckdb")),
            readLines("third_party/htslib/VERSION", n = 1L), cpu,
            "VCF/BCF: 1; BAM/CRAM: same as worker_count (1, 2, 4)", "0",
            "5 fresh processes per format and worker count; one untimed warm materialization")
)
knitr::kable(identity)

artifact_identity <- do.call(rbind, lapply(names(inputs), function(format) {
  receipt <- receipt_fields[[format]]
  data.frame(
    format = format,
    release = receipt[["release"]],
    records = as.numeric(receipt[["records"]]),
    samples = as.numeric(receipt[["samples"]]),
    bytes = file.info(inputs[[format]])$size,
    stringsAsFactors = FALSE
  )
}))
rownames(artifact_identity) <- NULL
knitr::kable(artifact_identity)

alignment_artifact_identity <- data.frame(
  format = c("BAM", "CRAM"),
  release = alignment_receipt_fields$BAM[["release"]],
  source_reads = 17000,
  panel_sites = 17000,
  bytes = file.info(alignment_inputs[c("BAM", "CRAM")])$size,
  index_bytes = file.info(alignment_inputs[c("bam_bai", "cram_crai")])$size,
  stringsAsFactors = FALSE
)
knitr::kable(alignment_artifact_identity)

panel_sql <- paste(
  "CREATE TEMP TABLE panel AS",
  "WITH candidate_sites AS (",
  "SELECT CHROM AS region, POS AS position, REF, ALT[1] AS ALT",
  "FROM read_bcf(%s, scan_mode := 'sequential', decompression_threads := 0)",
  "WHERE len(REF) = 1 AND len(ALT) = 1 AND len(ALT[1]) = 1",
  "AND REF <> ALT[1] AND REF IN ('A','C','G','T')",
  "AND ALT[1] IN ('A','C','G','T')",
  "), unique_sites AS (",
  "SELECT *, row_number() OVER (PARTITION BY region, position ORDER BY REF, ALT) AS occurrence",
  "FROM candidate_sites",
  "), selected AS (",
  "SELECT * FROM unique_sites WHERE occurrence = 1",
  "ORDER BY region, position, REF, ALT LIMIT 17000",
  ") SELECT 'GRCh38'::VARCHAR AS assembly,",
  "row_number() OVER (ORDER BY region, position, REF, ALT) - 1 AS site_index,",
  "region, position, least(REF, ALT) AS allele_a, greatest(REF, ALT) AS allele_b",
  "FROM selected"
)

measure_one <- function(extension, input, panel_source, format, repetition,
                        panel_sql, threads) {
  con <- DBI::dbConnect(duckdb::duckdb(
    config = list(allow_unsigned_extensions = "true")))
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE))
  quote <- function(value) as.character(DBI::dbQuoteString(con, value))
  DBI::dbExecute(con, paste("LOAD", quote(extension)))
  DBI::dbExecute(con, sprintf("SET threads=%d", threads))
  DBI::dbExecute(con, "SET memory_limit='4GB'")
  DBI::dbExecute(con, sprintf(panel_sql, quote(panel_source)))
  panel <- DBI::dbGetQuery(con, paste(
    "SELECT count(*) AS sites, min(site_index) AS first_site,",
    "max(site_index) AS last_site, count(DISTINCT (region, position)) AS positions,",
    "duckhts_somalier_panel_sha256('panel') AS panel_sha256 FROM panel GROUP BY ALL"))
  if (panel$sites[[1L]] != 17000 || panel$first_site[[1L]] != 0 ||
      panel$last_site[[1L]] != 16999 || panel$positions[[1L]] != 17000) {
    stop("derived panel contract changed")
  }
  sql <- sprintf(paste(
    "CREATE OR REPLACE TEMP TABLE extracted AS",
    "SELECT * FROM duckhts_somalier_vcf_counts(%s, 'panel')"), quote(input))
  DBI::dbExecute(con, sql)
  elapsed <- unname(system.time(DBI::dbExecute(con, sql))[["elapsed"]])
  counts <- DBI::dbGetQuery(con, paste(
    "SELECT count(*) AS output_rows, count(DISTINCT sample_id) AS output_samples,",
    "count(DISTINCT site_index) AS output_sites,",
    "count(*) FILTER (WHERE status = 'measured') AS measured_rows,",
    "count(*) FILTER (WHERE status <> 'measured') AS unavailable_rows",
    "FROM extracted"))
  if (counts$output_rows[[1L]] != 17000 || counts$output_samples[[1L]] != 1 ||
      counts$output_sites[[1L]] != 17000 || counts$measured_rows[[1L]] != 17000 ||
      counts$unavailable_rows[[1L]] != 0) {
    stop("extraction denominator or status changed")
  }
  status <- readLines("/proc/self/status", warn = FALSE)
  high_water <- status[grepl("^VmHWM:[[:space:]]", status)]
  if (length(high_water) != 1L) stop("Linux VmHWM is unavailable")
  peak_rss_kib <- as.numeric(sub(
    "^VmHWM:[[:space:]]*([0-9]+)[[:space:]]+kB$", "\\1", high_water))
  if (!is.finite(peak_rss_kib)) stop("could not parse Linux VmHWM")
  data.frame(format = format, repetition = repetition, elapsed_seconds = elapsed,
             peak_rss_mib = peak_rss_kib / 1024,
             input_records = 319349, panel_sites = panel$sites[[1L]],
             output_samples = counts$output_samples[[1L]],
             output_rows = counts$output_rows[[1L]],
             measured_rows = counts$measured_rows[[1L]],
             unavailable_rows = counts$unavailable_rows[[1L]],
             panel_sha256 = panel$panel_sha256[[1L]], stringsAsFactors = FALSE)
}

threads <- 1L
repetitions <- 5L
measurements <- list()
for (format in names(inputs)) {
  for (repetition in seq_len(repetitions)) {
    measurements[[length(measurements) + 1L]] <- callr::r(
      measure_one,
      args = list(extension = extension, input = inputs[[format]],
                  panel_source = inputs[["VCF"]], format = format,
                  repetition = repetition, panel_sql = panel_sql,
                  threads = threads),
      show = FALSE
    )
  }
}
measurements <- do.call(rbind, measurements)
if (length(unique(measurements$panel_sha256)) != 1L) {
  stop("panel identity changed between processes")
}

summary <- do.call(rbind, lapply(split(measurements, measurements$format), function(rows) {
  data.frame(
    format = rows$format[[1L]],
    input_records = rows$input_records[[1L]],
    panel_sites = rows$panel_sites[[1L]],
    output_samples = rows$output_samples[[1L]],
    output_rows = rows$output_rows[[1L]],
    measured_rows = rows$measured_rows[[1L]],
    unavailable_rows = rows$unavailable_rows[[1L]],
    threads = threads,
    median_seconds = median(rows$elapsed_seconds),
    minimum_seconds = min(rows$elapsed_seconds),
    maximum_seconds = max(rows$elapsed_seconds),
    median_peak_rss_mib = median(rows$peak_rss_mib),
    maximum_peak_rss_mib = max(rows$peak_rss_mib),
    input_records_per_second = rows$input_records[[1L]] /
      median(rows$elapsed_seconds),
    output_rows_per_second = rows$output_rows[[1L]] /
      median(rows$elapsed_seconds),
    stringsAsFactors = FALSE
  )
}))
rownames(summary) <- NULL
knitr::kable(summary, digits = 3)

validation <- callr::r(function(extension, inputs, panel_sql) {
  con <- DBI::dbConnect(duckdb::duckdb(
    config = list(allow_unsigned_extensions = "true")))
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE))
  quote <- function(value) as.character(DBI::dbQuoteString(con, value))
  DBI::dbExecute(con, paste("LOAD", quote(extension)))
  DBI::dbExecute(con, "SET threads=1")
  DBI::dbExecute(con, sprintf(panel_sql, quote(inputs[["VCF"]])))
  for (format in names(inputs)) {
    DBI::dbExecute(con, sprintf(
      "CREATE TEMP TABLE %s_evidence AS SELECT * FROM duckhts_somalier_vcf_counts(%s, 'panel')",
      tolower(format), quote(inputs[[format]])))
  }
  differences <- DBI::dbGetQuery(con, paste(
    "SELECT count(*) AS n FROM (",
    "(SELECT * EXCLUDE(source_path) FROM vcf_evidence",
    " EXCEPT ALL SELECT * EXCLUDE(source_path) FROM bcf_evidence)",
    "UNION ALL",
    "(SELECT * EXCLUDE(source_path) FROM bcf_evidence",
    " EXCEPT ALL SELECT * EXCLUDE(source_path) FROM vcf_evidence))"))$n[[1L]]
  data.frame(comparison = "VCF.gz versus BCF complete typed rows",
             left_rows = DBI::dbGetQuery(con, "SELECT count(*) n FROM vcf_evidence")$n[[1L]],
             right_rows = DBI::dbGetQuery(con, "SELECT count(*) n FROM bcf_evidence")$n[[1L]],
             symmetric_difference_rows = differences)
}, args = list(extension = extension, inputs = inputs, panel_sql = panel_sql), show = FALSE)
if (validation$symmetric_difference_rows[[1L]] != 0) {
  stop("VCF and BCF extracted relations differ")
}
knitr::kable(validation)

worker_capability <- callr::r(function(extension) {
  con <- DBI::dbConnect(duckdb::duckdb(
    config = list(allow_unsigned_extensions = "true")))
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE))
  quote <- function(value) as.character(DBI::dbQuoteString(con, value))
  DBI::dbExecute(con, paste("LOAD", quote(extension)))
  functions <- DBI::dbGetQuery(con, paste(
    "SELECT parameters FROM duckdb_functions()",
    "WHERE function_name = 'duckhts_somalier_bam_counts'"
  ))
  nrow(functions) == 1L && "worker_count" %in% functions$parameters[[1L]]
}, args = list(extension = extension), show = FALSE)
if (!isTRUE(worker_capability)) {
  stop(paste(
    "candidate duckhts_somalier_bam_counts() does not expose worker_count;",
    "the required 1/2/4-worker benchmark cannot run"
  ))
}

measure_alignment_one <- function(extension, source, index, reference,
                                  reference_index, panel, format,
                                  repetition, worker_count) {
  con <- DBI::dbConnect(duckdb::duckdb(
    config = list(allow_unsigned_extensions = "true")))
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE))
  quote <- function(value) as.character(DBI::dbQuoteString(con, value))
  DBI::dbExecute(con, paste("LOAD", quote(extension)))
  DBI::dbExecute(con, sprintf("SET threads=%d", worker_count))
  DBI::dbExecute(con, "SET memory_limit='4GB'")
  panel_counts <- DBI::dbGetQuery(con, paste(
    "SELECT count(*) AS sites, count(DISTINCT (region, position)) AS positions,",
    "min(site_index) AS first_site, max(site_index) AS last_site FROM read_parquet(",
    quote(panel), ")"
  ))
  if (panel_counts$sites[[1L]] != 17000 ||
      panel_counts$positions[[1L]] != 17000 ||
      panel_counts$first_site[[1L]] != 0 || panel_counts$last_site[[1L]] != 16999) {
    stop("synthetic alignment panel contract changed")
  }
  sql <- sprintf(paste(
    "CREATE OR REPLACE TEMP TABLE extracted AS SELECT * FROM",
    "duckhts_somalier_bam_counts(%s, NULL, 'synthetic', %s,",
    "panel_parquet := %s, index_path := %s, reference_index_path := %s,",
    "decompression_threads := 0, worker_count := %d)"
  ), quote(source), quote(reference), quote(panel), quote(index),
  quote(reference_index), worker_count)
  DBI::dbExecute(con, sql)
  elapsed <- unname(system.time(DBI::dbExecute(con, sql))[["elapsed"]])
  counts <- DBI::dbGetQuery(con, paste(
    "SELECT count(*) AS output_rows, count(DISTINCT sample_id) AS output_samples,",
    "count(DISTINCT site_index) AS output_sites,",
    "count(*) FILTER (WHERE status = 'measured') AS measured_rows,",
    "count(*) FILTER (WHERE status <> 'measured') AS unavailable_rows,",
    "sum(a) AS a, sum(b) AS b, sum(other) AS other, sum(a + b + other) AS observed_bases",
    "FROM extracted"
  ))
  if (counts$output_rows[[1L]] != 17000 || counts$output_samples[[1L]] != 1 ||
      counts$output_sites[[1L]] != 17000 || counts$measured_rows[[1L]] != 17000 ||
      counts$unavailable_rows[[1L]] != 0 || counts$a[[1L]] != 8500 ||
      counts$b[[1L]] != 8500 || counts$other[[1L]] != 0 ||
      counts$observed_bases[[1L]] != 17000) {
    stop("alignment extraction denominator, status, or count contract changed")
  }
  status <- readLines("/proc/self/status", warn = FALSE)
  high_water <- status[grepl("^VmHWM:[[:space:]]", status)]
  if (length(high_water) != 1L) stop("Linux VmHWM is unavailable")
  peak_rss_kib <- as.numeric(sub(
    "^VmHWM:[[:space:]]*([0-9]+)[[:space:]]+kB$", "\\1", high_water))
  if (!is.finite(peak_rss_kib)) stop("could not parse Linux VmHWM")
  data.frame(
    format = format, repetition = repetition, worker_count = worker_count,
    elapsed_seconds = elapsed, peak_rss_mib = peak_rss_kib / 1024,
    source_reads = 17000, panel_sites = panel_counts$sites[[1L]],
    output_samples = counts$output_samples[[1L]],
    output_rows = counts$output_rows[[1L]],
    measured_rows = counts$measured_rows[[1L]],
    unavailable_rows = counts$unavailable_rows[[1L]],
    observed_bases = counts$observed_bases[[1L]], stringsAsFactors = FALSE
  )
}

worker_counts <- c(1L, 2L, 4L)
alignment_measurements <- list()
for (format in c("BAM", "CRAM")) {
  index_name <- if (format == "BAM") "bam_bai" else "cram_crai"
  for (worker_count in worker_counts) {
    for (repetition in seq_len(repetitions)) {
      alignment_measurements[[length(alignment_measurements) + 1L]] <- callr::r(
        measure_alignment_one,
        args = list(
          extension = extension, source = alignment_inputs[[format]],
          index = alignment_inputs[[index_name]],
          reference = alignment_inputs[["reference"]],
          reference_index = alignment_inputs[["reference_fai"]],
          panel = alignment_inputs[["panel"]], format = format,
          repetition = repetition, worker_count = worker_count
        ),
        show = FALSE
      )
    }
  }
}
alignment_measurements <- do.call(rbind, alignment_measurements)
alignment_groups <- split(
  alignment_measurements,
  interaction(alignment_measurements$format,
              alignment_measurements$worker_count, drop = TRUE)
)
alignment_summary <- do.call(rbind, lapply(alignment_groups, function(rows) {
  data.frame(
    format = rows$format[[1L]], source_reads = rows$source_reads[[1L]],
    panel_sites = rows$panel_sites[[1L]],
    output_samples = rows$output_samples[[1L]], output_rows = rows$output_rows[[1L]],
    measured_rows = rows$measured_rows[[1L]],
    unavailable_rows = rows$unavailable_rows[[1L]],
    worker_count = rows$worker_count[[1L]],
    median_seconds = median(rows$elapsed_seconds),
    minimum_seconds = min(rows$elapsed_seconds),
    maximum_seconds = max(rows$elapsed_seconds),
    median_peak_rss_mib = median(rows$peak_rss_mib),
    maximum_peak_rss_mib = max(rows$peak_rss_mib),
    source_reads_per_second = rows$source_reads[[1L]] /
      median(rows$elapsed_seconds),
    output_rows_per_second = rows$output_rows[[1L]] /
      median(rows$elapsed_seconds), stringsAsFactors = FALSE
  )
}))
alignment_summary <- alignment_summary[
  order(alignment_summary$format, alignment_summary$worker_count), , drop = FALSE
]
rownames(alignment_summary) <- NULL
knitr::kable(alignment_summary, digits = 3)

alignment_validation <- callr::r(function(extension, inputs) {
  con <- DBI::dbConnect(duckdb::duckdb(
    config = list(allow_unsigned_extensions = "true")))
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE))
  quote <- function(value) as.character(DBI::dbQuoteString(con, value))
  DBI::dbExecute(con, paste("LOAD", quote(extension)))
  DBI::dbExecute(con, "SET threads=4")
  for (format in c("BAM", "CRAM")) {
    index_name <- if (format == "BAM") "bam_bai" else "cram_crai"
    for (worker_count in c(1L, 2L, 4L)) {
      DBI::dbExecute(con, sprintf(paste(
        "CREATE TEMP TABLE %s_w%d AS SELECT * FROM",
        "duckhts_somalier_bam_counts(%s, NULL, 'synthetic', %s,",
        "panel_parquet := %s, index_path := %s, reference_index_path := %s,",
        "decompression_threads := 0, worker_count := %d)"
      ), tolower(format), worker_count, quote(inputs[[format]]),
      quote(inputs[["reference"]]), quote(inputs[["panel"]]),
      quote(inputs[[index_name]]),
      quote(inputs[["reference_fai"]]), worker_count))
    }
  }
  compare <- function(left, right, comparison, excluded_columns = "none") {
    projection <- if (excluded_columns == "none") "*" else "* EXCLUDE(source_path)"
    differences <- DBI::dbGetQuery(con, sprintf(paste(
      "SELECT count(*) AS n FROM (",
      "(SELECT %s FROM %s EXCEPT ALL SELECT %s FROM %s)",
      "UNION ALL",
      "(SELECT %s FROM %s EXCEPT ALL SELECT %s FROM %s))"
    ), projection, left, projection, right,
    projection, right, projection, left))$n[[1L]]
    data.frame(
      comparison = comparison, excluded_columns = excluded_columns,
      left_rows = DBI::dbGetQuery(
        con, paste("SELECT count(*) n FROM", left))$n[[1L]],
      right_rows = DBI::dbGetQuery(
        con, paste("SELECT count(*) n FROM", right))$n[[1L]],
      symmetric_difference_rows = differences, stringsAsFactors = FALSE
    )
  }
  validations <- list()
  for (format in c("BAM", "CRAM")) {
    for (worker_count in c(2L, 4L)) {
      validations[[length(validations) + 1L]] <- compare(
        paste0(tolower(format), "_w1"),
        paste0(tolower(format), "_w", worker_count),
        sprintf("%s worker_count 1 versus %d", format, worker_count)
      )
    }
  }
  validations[[length(validations) + 1L]] <- compare(
    "bam_w4", "cram_w4", "BAM versus CRAM at worker_count 4", "source_path"
  )
  do.call(rbind, validations)
}, args = list(extension = extension, inputs = alignment_inputs), show = FALSE)
if (any(alignment_validation$symmetric_difference_rows != 0)) {
  stop("alignment extraction relations differ across formats or worker counts")
}
knitr::kable(alignment_validation)
```

## Result

At one DuckDB thread, median complete materialization was
`r sprintf("%.3f", summary$median_seconds[summary$format == "VCF"])` seconds from
VCF.gz and
`r sprintf("%.3f", summary$median_seconds[summary$format == "BCF"])` seconds from
BCF. Both runs consumed 319,349 physical records and emitted exactly 17,000
measured rows for one sample. Median whole-process peak RSS was
`r sprintf("%.1f", summary$median_peak_rss_mib[summary$format == "VCF"])` MiB
for VCF.gz and
`r sprintf("%.1f", summary$median_peak_rss_mib[summary$format == "BCF"])` MiB
for BCF. The exact bidirectional `EXCEPT ALL` comparison,
excluding only the intentionally different source path, found
`r validation$symmetric_difference_rows[[1L]]` differing rows.

This is the first recorded workload for the extraction surface, so there is no
identical historical baseline and no before/after speedup claim. The comparison
shows encoding and worker-count measurements within the same candidate build.

For the synthetic aligned-read workload, the measured BAM medians for worker
counts 1, 2, and 4 were
`r paste(sprintf("%.3f", alignment_summary$median_seconds[alignment_summary$format == "BAM"]), collapse = ", ")`
seconds; the corresponding CRAM medians were
`r paste(sprintf("%.3f", alignment_summary$median_seconds[alignment_summary$format == "CRAM"]), collapse = ", ")`
seconds. Every timed materialization consumed 17,000 source reads across 17,000
panel sites and emitted 17,000 measured rows for one sample. Whole-process
median peak RSS for BAM was
`r paste(sprintf("%.1f", alignment_summary$median_peak_rss_mib[alignment_summary$format == "BAM"]), collapse = ", ")`
MiB at worker counts 1, 2, and 4; CRAM used
`r paste(sprintf("%.1f", alignment_summary$median_peak_rss_mib[alignment_summary$format == "CRAM"]), collapse = ", ")`
MiB. The four exact worker-count comparisons, which excluded no columns,
found at most
`r max(alignment_validation$symmetric_difference_rows[alignment_validation$excluded_columns == "none"])`
differing rows. The BAM-versus-CRAM comparison at four workers excluded only
the intentionally different source path and found
`r alignment_validation$symmetric_difference_rows[alignment_validation$excluded_columns == "source_path"]`
differing rows.

These measurements do not establish a general scaling result. They describe
this small synthetic sparse-panel workload on the recorded machine and candidate
binary. The report does not measure a multi-sample cohort, remote I/O, output
persisted to Parquet, whole-genome alignments, or downstream sketch,
relatedness, CHARR, or contamination computation.
