---
title: "Record-major genotypes and carrier expansion on an HPRC cohort"
output:
  github_document:
    html_preview: false
---

The registered HPRC v2.0 GRCh38 cohort region `chr22:20000000-21000000`
retains every source sample and allele. This is not the consequence-only,
split-allele corpus. Stage explicitly with
`duckhtsbench::duckhts_bench_stage_genotypes()` and
`duckhtsbench::duckhts_bench_stage_genotype_phase_set()` before rendering;
rendering is network-free. BCF and VCF.gz contain the same complete records,
verified by the staging test and full HTSlib text comparison.

This compares `read_geno()` with the canonical `read_bcf(..., tidy_format:=true)`
GT projection. Full cohorts, the
first eight header samples, sparse non-reference calls, and downstream typed
carrier expansion have separate denominators. The input has GT but no PS;
these timings do not measure PS decoding, which SQL/R/native fixtures test.

`benchmark_genotypes_format_shared.md` and `benchmark_genotypes_scalar_counts.md`
are revision-specific snapshots of this Rmd. Preserve their measured revisions;
render another snapshot from the repository root with a new report name:

```sh
BENCHMARK_RMD=benchmark_genotypes.Rmd \
BENCHMARK_REPORT=benchmark_genotypes_review.md taskset -c 2 make bench-snapshot
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
knitr::opts_knit$set(root.dir=normalizePath(".."))
```

```{r measurement, include=FALSE}
library(DBI)
library(duckhtsbench)
source("benchmarks/genotype_reader_run.R")
extension <- normalizePath(Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension"), mustWork=TRUE)
stopifnot(system2("git", c("diff", "--quiet", "HEAD", "--", "src")) == 0L)
revision <- system2("git", c("rev-parse", "HEAD"), stdout=TRUE)
src_tree <- system2("git", c("rev-parse", "HEAD:src"), stdout=TRUE)
paths <- c(VCF = duckhts_bench_artifact_path("geno_hprc_vcfgz"), BCF = duckhts_bench_artifact_path("geno_hprc_bcf"))
identities <- list()
for (i in seq_along(paths)) {
  id <- c("geno_hprc_vcfgz", "geno_hprc_bcf")[[i]]
  duckhts_bench_validate_identity(id, paths[[i]])
  receipt <- read.delim(paste0(paths[[i]], ".provenance.tsv"))
  values <- setNames(receipt$value, receipt$field)
  stopifnot(identical(unname(tools::md5sum(paths[[i]])), values[["observed_md5"]]))
  identities[[i]] <- data.frame(format=names(paths)[i], artifact=id,
                                bytes=file.info(paths[[i]])$size, md5=values[["observed_md5"]])
}
identities <- do.call(rbind, identities)
bcftools <- function(args) {
  result <- system2("bcftools", shQuote(args), stdout=TRUE)
  stopifnot(is.null(attr(result,"status")))
  result
}
samples <- bcftools(c("query","-l",paths[[1]]))
header <- bcftools(c("view","-h",paths[[1]]))
stopifnot("##fileformat=VCFv4.2" %in% header,
          sum(startsWith(header,"##FORMAT=")) == 1L,
          any(startsWith(header,"##FORMAT=<ID=GT,")), length(samples) >= 8L)
gt_lines <- bcftools(c("query","-f","[%GT\t]\n",paths[[1]]))
input_records <- length(gt_lines)
# Independent HTSlib GT strings supply all workload denominators. No optimized
# reader count path, dropped records or changed oracle is used for a rate.
denominators <- lapply(c("full","selected","sparse","carriers"), function(workload) {
  selected <- if (workload == "selected") seq_len(8L) else seq_along(samples)
  sums <- vapply(strsplit(gt_lines,"\t",fixed=TRUE), function(gt) {
    stopifnot(length(gt) == length(samples))
    gt <- gt[selected]
    slots <- nchar(gsub("[^|/]","",gt)) + 1L
    nonref <- grepl("(^|[|/])[1-9][0-9]*($|[|/])",gt)
    alt <- lengths(regmatches(gt, gregexpr("(^|[|/])[1-9][0-9]*",gt)))
    c(calls=length(gt), slots=sum(slots), nonref=sum(nonref),
      sparse_slots=sum(slots[nonref]), alt_slots=sum(alt))
  }, numeric(5))
  total <- rowSums(sums)
  data.frame(workload, input_records, selected_samples=length(selected),
             selected_calls=total[["calls"]], selected_slots=total[["slots"]],
             non_reference_calls=total[["nonref"]], sparse_slots=total[["sparse_slots"]],
             alt_slots=total[["alt_slots"]])
})
denominators <- do.call(rbind,denominators)
rm(gt_lines)
repetitions <- as.integer(Sys.getenv("GENO_BENCH_RUNS","3"))
stopifnot(repetitions > 0L)
directory <- tempfile("duckhts-genotype-report-")
dir.create(directory)
results <- comparisons <- list()
con <- dbConnect(duckdb::duckdb())
dbExecute(con,"SET threads=1")
dbExecute(con,"SET memory_limit='8GB'")
for (format in names(paths)) for (workload in denominators$workload) {
  expected <- denominators[denominators$workload==workload,]
  snapshots <- c()
  for (repetition in seq_len(repetitions)) {
    readers <- if (repetition %% 2L) c("read_bcf","read_geno") else c("read_geno","read_bcf")
    for (reader in readers) {
      snapshot <- if (repetition==1L) file.path(directory,paste(format,workload,reader,"parquet",sep=".")) else ""
      if (nzchar(snapshot)) snapshots[reader] <- snapshot
      observed <- callr::r(genotype_reader_run, args=list(extension=extension, input=paths[[format]],
        reader=reader, workload=workload,
        selector=if(workload=="selected") paste(samples[seq_len(8L)],collapse=",") else "",
        sample_names=samples, snapshot=snapshot, threads=1L))
      if (workload=="carriers") stopifnot(observed$output_rows==expected$alt_slots)
      else {
        stopifnot(observed$output_calls==if(workload=="sparse") expected$non_reference_calls else expected$selected_calls,
                  observed$output_slots==if(workload=="sparse") expected$sparse_slots else expected$selected_slots,
                  observed$output_rows==if(reader=="read_geno") input_records else observed$output_calls)
      }
      results[[length(results)+1L]] <- cbind(format, repetition, observed)
    }
  }
  reference <- paste0("read_parquet(",dbQuoteString(con,snapshots[["read_bcf"]]),")")
  candidate <- paste0("read_parquet(",dbQuoteString(con,snapshots[["read_geno"]]),")")
  sql <- sprintf(paste("SELECT count(*) AS n FROM ((SELECT * FROM %s EXCEPT ALL SELECT * FROM %s)",
    "UNION ALL (SELECT * FROM %s EXCEPT ALL SELECT * FROM %s))"),reference,candidate,candidate,reference)
  difference <- as.numeric(dbGetQuery(con,sql)$n)
  stopifnot(difference==0)
  comparisons[[length(comparisons)+1L]] <- data.frame(format,workload,different_typed_rows=difference)
  unlink(snapshots)
}
dbDisconnect(con,shutdown=TRUE)
unlink(directory,recursive=TRUE)
results <- do.call(rbind,results)
comparisons <- do.call(rbind,comparisons)
for (key in split(results,interaction(results$format,results$reader,results$workload)))
  stopifnot(length(unique(key$checksum))==1L)
```

## Source and workload identity

```{r identity}
cat("Source revision: ", revision, "\nSource tree: ", src_tree, "\n", sep="")
cat("Extension MD5:",unname(tools::md5sum(extension)),"\n")
cat("R:",R.version.string,"; DuckDB:",as.character(packageVersion("duckdb")),"\n")
cat(system2("uname","-srmo",stdout=TRUE),"\n")
cat(system2("taskset",c("-pc",Sys.getpid()),stdout=TRUE),"\n")
cat(system2("sh",c("-c",shQuote("lscpu | sed -n '/Model name:/p'")),stdout=TRUE),"\n")
cat("DuckDB threads: 1; scan handles: 1; HTSlib decompression workers: 0; repetitions:",repetitions,"\n")
knitr::kable(identities)
knitr::kable(denominators)
```

## Materialization and carrier expansion

Timing includes binding, decoding, selected output vectors and `CREATE TABLE AS`.
Carrier timing also includes allele-slot expansion and ALT selection. Every
measurement uses a fresh R/DuckDB process; source data is warm in the OS cache.
Startup, independent counting, checksums, checkpoint, snapshots and exact output
comparison are outside timing. RSS is the process high-water mark immediately
after materialization. Materialized database bytes are measured after checkpoint
and include the tiny sample catalog and database overhead; they are physical
compressed storage, not vector-memory bytes. The fixed DuckDB memory limit is
8 GiB; these are not sealed/static-allocation measurements.

```{r timing}
summary <- aggregate(cbind(elapsed,peak_rss_kib,materialized_database_bytes) ~ format+reader+workload,results,median)
summary <- merge(summary,denominators,by="workload",sort=FALSE)
summary$records_per_second <- summary$input_records/summary$elapsed
summary$calls_per_second <- summary$selected_calls/summary$elapsed
summary$allele_slots_per_second <- summary$selected_slots/summary$elapsed
summary$input_bytes_per_second <- identities$bytes[match(summary$format,identities$format)]/summary$elapsed
knitr::kable(summary[c("format","reader","workload","elapsed","records_per_second","calls_per_second",
                      "allele_slots_per_second","input_bytes_per_second","peak_rss_kib","materialized_database_bytes")],digits=3)
knitr::kable(results[c("format","reader","workload","repetition","output_rows","output_calls","output_slots","elapsed","cpu")],digits=3)
```

## Complete output comparison

First-run results are normalized outside the timer to typed variant identity,
original-header sample index, every allele slot, every phase flag and phase set.
For the VCF 4.2 source, the comparison reconstructs the leading phase flag from
HTSlib's string representation; VCF 4.4 explicit-prefix semantics are tested in
the committed SQL/R witnesses, not inferred from this dataset. Every field and
duplicate contributes to `EXCEPT ALL` in both directions. Carrier comparisons
retain variant identity, sample, allele slot/index, selected ALT, phase and PS.
Each repetition separately checks the complete materialized-row checksum.

```{r agreement}
knitr::kable(comparisons)
```

The nearest earlier full-materialization report is
[the shared BCF scanner report](benchmark_bcf_shared_scan.md): a single-sample
GIAB workload, not an identical cohort baseline. This report compares the two
current interfaces, not pre-change and post-change revisions. The separately
rendered paired BCF regression workload measures changes to the existing reader.
These results do not establish universal fastest-reader performance, remote-I/O
throughput or multi-worker scaling. This HPRC section does not measure PS
throughput or phased annotation correctness.

```{r phase-set-measurement, include=FALSE}
phase_ids <- c(VCF="geno_giab_phased_chr1_vcfgz", BCF="geno_giab_phased_chr1_bcf")
phase_paths <- vapply(phase_ids, duckhts_bench_artifact_path, character(1L))
phase_identity <- list()
for (format in names(phase_paths)) {
  id <- phase_ids[[format]]
  path <- phase_paths[[format]]
  duckhts_bench_validate_identity(id, path)
  receipt_path <- paste0(path, ".provenance.tsv")
  stopifnot(file.exists(receipt_path))
  receipt <- read.delim(receipt_path, colClasses="character", check.names=FALSE)
  receipt <- setNames(receipt$value, receipt$field)
  stopifnot(all(c("source_artifact", "source_supplier_identity", "source_index_artifact",
                  "bcftools_version", "observed_sha256", "records", "samples", "calls",
                  "allele_slots", "nonnull_ps") %in% names(receipt)))
  observed_sha256 <- strsplit(system2("sha256sum", shQuote(path), stdout=TRUE),
                              "[[:space:]]+")[[1L]][[1L]]
  stopifnot(identical(observed_sha256, receipt[["observed_sha256"]]))
  phase_identity[[format]] <- data.frame(
    format, artifact=id, bytes=file.info(path)$size, sha256=observed_sha256,
    source=receipt[["source_artifact"]], stringsAsFactors=FALSE)
}
phase_identity <- do.call(rbind, phase_identity)
rownames(phase_identity) <- NULL

phase_oracle <- function(path) {
  samples <- bcftools(c("query", "-l", path))
  records <- bcftools(c("query", "-f", "%CHROM\\n", path))
  calls <- bcftools(c("query", "-f", "[%GT\\t%PS\\n]", path))
  stopifnot(length(samples) > 0L, length(calls) == length(records) * length(samples))
  fields <- strsplit(calls, "\t", fixed=TRUE)
  stopifnot(all(lengths(fields) == 2L))
  gt <- vapply(fields, `[[`, character(1L), 1L)
  ps <- vapply(fields, `[[`, character(1L), 2L)
  list(samples=samples, calls=calls, counts=c(
    records=length(records), samples=length(samples), calls=length(calls),
    allele_slots=sum(nchar(gsub("[^|/]", "", gt)) + 1L), nonnull_ps=sum(ps != ".")))
}
phase_oracles <- lapply(phase_paths, phase_oracle)
stopifnot(identical(phase_oracles[["VCF"]]$samples, phase_oracles[["BCF"]]$samples),
          identical(phase_oracles[["VCF"]]$calls, phase_oracles[["BCF"]]$calls),
          identical(phase_oracles[["VCF"]]$counts, phase_oracles[["BCF"]]$counts),
          phase_oracles[["VCF"]]$counts[["nonnull_ps"]] > 0)
phase_counts <- as.data.frame(as.list(phase_oracles[["VCF"]]$counts))
phase_counts[] <- lapply(phase_counts, as.numeric)

phase_repetitions <- as.integer(Sys.getenv("GENO_PS_RUNS", "5"))
stopifnot(length(phase_repetitions) == 1L, !is.na(phase_repetitions), phase_repetitions >= 5L)
phase_directory <- tempfile("duckhts-genotype-phase-set-report-")
dir.create(phase_directory)
phase_results <- list()
phase_snapshots <- setNames(vector("list", length(phase_paths)), names(phase_paths))
for (format in names(phase_paths)) {
  for (repetition in seq_len(phase_repetitions)) {
    readers <- if (repetition %% 2L) c("read_bcf", "read_geno") else c("read_geno", "read_bcf")
    for (reader in readers) {
      snapshot <- if (repetition == 1L) {
        file.path(phase_directory, paste(format, reader, "parquet", sep="."))
      } else ""
      if (nzchar(snapshot)) phase_snapshots[[format]][[reader]] <- snapshot
      observed <- callr::r(genotype_reader_run, args=list(
        extension=extension, input=phase_paths[[format]], reader=reader, workload="full",
        selector="", sample_names=phase_oracles[[format]]$samples, snapshot=snapshot,
        threads=1L, phase_set_field=TRUE))
      expected <- phase_oracles[[format]]$counts
      stopifnot(observed$output_rows == if (reader == "read_geno") expected[["records"]] else expected[["calls"]],
                observed$output_calls == expected[["calls"]],
                observed$output_slots == expected[["allele_slots"]],
                observed$output_ps_values == expected[["nonnull_ps"]])
      phase_results[[length(phase_results) + 1L]] <- cbind(format, repetition, observed)
    }
  }
}
phase_results <- do.call(rbind, phase_results)
for (rows in split(phase_results, interaction(phase_results$format, phase_results$reader)))
  stopifnot(length(unique(rows$checksum)) == 1L)

phase_con <- dbConnect(duckdb::duckdb())
dbExecute(phase_con, "SET threads=1")
phase_comparisons <- list()
phase_table <- function(path) paste0("read_parquet(", dbQuoteString(phase_con, path), ")")
for (format in names(phase_paths)) {
  reference <- paste("SELECT * FROM", phase_table(phase_snapshots[[format]][["read_bcf"]]))
  candidate <- paste("SELECT * FROM", phase_table(phase_snapshots[[format]][["read_geno"]]))
  differences <- genotype_reader_difference(phase_con, reference, candidate)
  stopifnot(differences == 0)
  phase_comparisons[[length(phase_comparisons) + 1L]] <- data.frame(
    format, comparison="read_bcf versus read_geno", differences)
}
candidate <- paste("SELECT * FROM", phase_table(phase_snapshots[["VCF"]][["read_geno"]]))
identity_columns <- "CHROM,POS,ID,REF,ALT,sample_index"
controls <- c(
  duplicate_multiplicity=genotype_reader_difference(phase_con,
    paste0("SELECT * FROM (", candidate, ") UNION ALL (SELECT * FROM (", candidate, ") LIMIT 1)"),
    candidate),
  null_phase_set=genotype_reader_difference(phase_con, candidate,
    paste0("SELECT ", identity_columns, ",alleles,phase_before,NULL::BIGINT AS phase_set FROM (", candidate, ")")),
  changed_phase_set=genotype_reader_difference(phase_con, candidate,
    paste0("SELECT ", identity_columns, ",alleles,phase_before,",
           "CASE WHEN phase_set IS NULL THEN NULL ELSE phase_set+1 END AS phase_set FROM (", candidate, ")")),
  removed_allele_slot=genotype_reader_difference(phase_con, candidate,
    paste0("SELECT ", identity_columns, ",list_slice(alleles,1,len(alleles)-1) AS alleles,",
           "phase_before,phase_set FROM (", candidate, ")")),
  changed_phase_bit=genotype_reader_difference(phase_con, candidate,
    paste0("SELECT ", identity_columns, ",alleles,",
           "list_concat([NOT phase_before[1]],list_slice(phase_before,2,len(phase_before))) AS phase_before,",
           "phase_set FROM (", candidate, ")"))
)
stopifnot(all(controls > 0))
dbDisconnect(phase_con, shutdown=TRUE)
unlink(phase_directory, recursive=TRUE)

phase_summary <- aggregate(
  cbind(elapsed,cpu,peak_rss_kib,materialized_database_bytes) ~ format+reader,
  phase_results, median)
phase_summary$records_per_second <- phase_counts$records / phase_summary$elapsed
phase_summary$calls_per_second <- phase_counts$calls / phase_summary$elapsed
phase_summary$allele_slots_per_second <- phase_counts$allele_slots / phase_summary$elapsed
phase_summary$phase_sets_per_second <- phase_counts$nonnull_ps / phase_summary$elapsed
phase_summary$input_bytes_per_second <- phase_identity$bytes[
  match(phase_summary$format, phase_identity$format)] / phase_summary$elapsed
```

## Non-null FORMAT/PS materialization

The pinned GIAB HG002 v4.2.1 phased benchmark supplies an integer `FORMAT/PS`
oracle that the existing HPRC cohort and unphased GIAB input do not. Staging
retains every chr1 record and the original single sample, then encodes those
same records as VCF.gz and BCF. It does not generate phase sets or filter calls.

Stage the registered source and derived inputs explicitly before rendering:

```sh
Rscript -e 'duckhtsbench::duckhts_bench_stage_genotype_phase_set()'
```

```{r phase-set-identity}
cat("Source revision: ", revision, "\nSource tree: ", src_tree, "\n", sep="")
cat("Extension MD5:",unname(tools::md5sum(extension)),"\n")
cat("R:",R.version.string,"; DuckDB:",as.character(packageVersion("duckdb")),"\n")
cat(system2("uname","-srmo",stdout=TRUE),"\n")
cat(system2("taskset",c("-pc",Sys.getpid()),stdout=TRUE),"\n")
cat(system2("sh",c("-c",shQuote("lscpu | sed -n '/Model name:/p'")),stdout=TRUE),"\n")
cat("DuckDB threads: 1; scan handles: 1; HTSlib decompression workers: 0; repetitions:",
    phase_repetitions,"\n")
knitr::kable(phase_identity)
knitr::kable(data.frame(denominator=names(phase_oracles[["VCF"]]$counts),
                        count=as.numeric(phase_oracles[["VCF"]]$counts)))
```

The independent `bcftools query` oracle reads both encodings and requires the
complete GT/PS call streams to be identical. Every physical record, call,
allele slot and non-null phase set contributes to the denominator.

```{r phase-set-timing}
knitr::kable(phase_summary[c("format","reader","elapsed","cpu","records_per_second",
  "calls_per_second","allele_slots_per_second","phase_sets_per_second",
  "input_bytes_per_second","peak_rss_kib","materialized_database_bytes")], digits=3)
knitr::kable(phase_results[c("format","reader","repetition","output_rows","output_calls",
  "output_slots","output_ps_values","elapsed","cpu","peak_rss_kib")], digits=3)
```

Each timer includes sequential bind, read/decompression, typed GT/PS decoding,
vector writing and CTAS in a fresh process. Checksums, denominators, Parquet
snapshots and comparisons are outside timing. Reader order alternates between
repetitions. Peak RSS is the process high-water mark after materialization.

```{r phase-set-agreement}
knitr::kable(do.call(rbind, phase_comparisons))
knitr::kable(data.frame(control=names(controls), rejected_differences=as.numeric(controls)))
```

The normalized comparison retains variant identity, original-header sample
ordinal, every allele and phase slot, and nullable PS. `EXCEPT ALL` runs in both
directions, so duplicates retain multiplicity. Independent controls prove that
a changed duplicate multiplicity, NULL or changed PS, removed allele slot and
changed phase bit are rejected. The one-sample source does not support a
meaningful selected-sample or sparse comparison; those remain measured by the
HPRC cohort above.
