---
title: "CIGAR aligned blocks benchmark"
output:
  github_document:
    html_preview: false
    pandoc_args: ["--wrap=preserve", "--columns=400"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

This report measures `cigar_aligned_blocks(CIGAR, POS)` against the two ways
the same geometry was available before it existed: reading the binary CIGAR
and doing nothing with it, which is the denominator, and deriving the blocks in
SQL with a `list_slice` prefix sum inside a lambda, which is quadratic in the
op count of each read. All three are timed on the same host in one render,
through one in-memory DuckDB connection, over two real alignments staged
through the duckhtsbench registry:

- the `ont-ecoli-k12` workload: ENA run `ERR14686255`, 25,950 MinION reads of
  an *Escherichia coli* K-12 MG1655 derivative, aligned by `minimap2 -x map-ont`
  to the NCBI RefSeq assembly `GCF_000005845.2` and sorted by `samtools` at
  staging time. The genome is one contig, so `read_bam` scans it on one thread.
- the `riker-wgs` workload: the 1000 Genomes high-coverage sample HG00188
  (`ERR3240174`) transcoded from CRAM to BAM at staging time, read through the
  `chr22` region of its index. A region scan is one `read_bam` thread by
  construction, and it keeps the short-read condition to a few million reads.

The SQL derivation and the scalar must produce the same blocks: every timed
pass computes an order-independent fingerprint over every read's block lists,
and the render stops if the two derivations disagree on either input. That
makes each render a whole-file check of the scalar against an independent
formulation, on top of the fixture oracle in `test/sql/cigar_aligned_blocks.test`.

There is no earlier checked-in benchmark of this function. The nearest report
on the short-read input is [`benchmark_riker_wgs.md`](benchmark_riker_wgs.md),
which times whole-file `read_bam` scans against Riker; it shares the registered
BAM but not the region or the projection, so it is context rather than a
comparison.

## Reproduction

Stage both inputs once, with network access only in this step. The long-read
staging needs `minimap2` and `samtools`; the short-read staging needs
`samtools` and downloads a 14 GB CRAM plus the GRCh38 reference:

``` sh
Rscript -e 'duckhtsbench::duckhts_bench_stage_ont_ecoli()'
Rscript -e 'duckhtsbench::duckhts_bench_stage_riker()'
```

`DUCKHTS_EXTENSION` selects the extension to measure (default:
`build/release/duckhts.duckdb_extension`). Render from a tracked-clean revision;
rendering resolves the staged paths through the registry and does not download:

``` sh
taskset -c 8-11 Rscript -e 'rmarkdown::render("benchmarks/benchmark_cigar_aligned_blocks.Rmd")'
```

Both inputs run on one `read_bam` thread, so the benchmark pins each pass to
the highest logical CPU in the affinity mask of the rendering process; wrap the
render in `taskset -c` to choose it. Input staging, extension loading,
connection setup, warm-up and result verification are excluded from timing.

```{r benchmark, echo=FALSE, message=FALSE}
suppressPackageStartupMessages({
  library(DBI)
  library(duckdb)
})

sql_string <- function(x) {
  sprintf("'%s'", gsub("'", "''", x, fixed = TRUE))
}

command_output <- function(command, args = character()) {
  paste(system2(command, args, stdout = TRUE, stderr = TRUE), collapse = " ")
}

receipt_field <- function(receipt, field) {
  fields <- utils::read.delim(receipt, colClasses = "character", quote = "", comment.char = "")
  value <- fields$value[fields$field == field]
  if (length(value) != 1L) stop("missing provenance field ", field, " in ", receipt)
  value
}

ont <- duckhtsbench::duckhts_bench_stage_ont_ecoli(fetch = FALSE)
riker_bam <- duckhtsbench::duckhts_bench_artifact_path("riker_hg00188_bam")
stopifnot(
  file.exists(riker_bam),
  file.exists(paste0(riker_bam, ".bai")),
  file.exists(paste0(riker_bam, ".provenance.tsv"))
)
inputs <- list(
  ont = list(
    label = "ONT E. coli K-12 (ERR14686255)",
    path = normalizePath(ont[["bam"]], mustWork = TRUE),
    region = NULL
  ),
  riker = list(
    label = "Illumina HG00188 chr22 (ERR3240174)",
    path = normalizePath(riker_bam, mustWork = TRUE),
    region = "chr22"
  )
)

extension <- normalizePath(Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension"),
                           mustWork = TRUE)
revision <- command_output("git", c("rev-parse", "--short=12", "HEAD"))
src_tree <- command_output("git", c("rev-parse", "HEAD:src"))
tracked_dirty <- length(system2(
  "git", c("status", "--porcelain", "--untracked-files=no"), stdout = TRUE
)) > 0L
if (tracked_dirty) {
  stop("Render the benchmark from a tracked-clean source revision")
}

con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
invisible(dbExecute(con, paste("LOAD", dbQuoteString(con, extension))))

scan_sql <- function(input) {
  region <- if (is.null(input$region)) "" else sprintf(", region := %s", sql_string(input$region))
  sprintf(
    "read_bam(%s, index_path := %s, cigar_representation := 'binary'%s)",
    sql_string(input$path), sql_string(paste0(input$path, ".bai")), region
  )
}

# The quadratic SQL derivation: for every op, slice the preceding ops and sum
# what they consume. Reference-consuming ops are M D N = X (0 2 3 7 8), query-
# consuming ops are M I S = X (0 1 4 7 8), and a block is an M, = or X op.
sql_blocks <- paste0(
  "list_transform(list_filter(list_transform(CIGAR, lambda c, i: {'kind': c & 15, ",
  "'v': POS + COALESCE(list_sum(list_transform(list_slice(CIGAR, 1, i - 1), ",
  "lambda p: CASE WHEN (p & 15) IN (0, 2, 3, 7, 8) THEN (p >> 4)::BIGINT ELSE 0 END)), 0)}), ",
  "lambda x: x.kind IN (0, 7, 8)), lambda x: x.v) AS ref_start, ",
  "list_transform(list_filter(list_transform(CIGAR, lambda c, i: {'kind': c & 15, ",
  "'v': COALESCE(list_sum(list_transform(list_slice(CIGAR, 1, i - 1), ",
  "lambda p: CASE WHEN (p & 15) IN (0, 1, 4, 7, 8) THEN (p >> 4)::BIGINT ELSE 0 END)), 0)}), ",
  "lambda x: x.kind IN (0, 7, 8)), lambda x: x.v) AS query_start, ",
  "list_transform(list_filter(CIGAR, lambda c: (c & 15) IN (0, 7, 8)), lambda c: (c >> 4)::BIGINT) AS width"
)
scalar_blocks <- paste0(
  "(b).ref_start AS ref_start, (b).query_start AS query_start, (b).width AS width"
)

# One row back to R. The fingerprint hashes every read's three block lists with
# its identity, so the SQL and scalar derivations must agree read for read to
# produce the same value; the plain scan hashes the packed CIGAR instead and
# reads the same rows.
workload_sql <- function(input, workload) {
  switch(
    workload,
    "read_bam only" = sprintf(
      paste0(
        "SELECT count(*)::UBIGINT AS reads, sum(len(CIGAR))::DOUBLE AS ops, 0::DOUBLE AS blocks, ",
        "hex(bit_xor(hash(QNAME, FLAG, POS, CIGAR))) AS fingerprint FROM %s"
      ),
      scan_sql(input)
    ),
    "blocks in SQL" = sprintf(
      paste0(
        "SELECT count(*)::UBIGINT AS reads, sum(ops)::DOUBLE AS ops, sum(len(width))::DOUBLE AS blocks, ",
        "hex(bit_xor(hash(QNAME, FLAG, ref_start, query_start, width))) AS fingerprint ",
        "FROM (SELECT QNAME, FLAG, len(CIGAR) AS ops, %s FROM %s)"
      ),
      sql_blocks, scan_sql(input)
    ),
    "cigar_aligned_blocks" = sprintf(
      paste0(
        "SELECT count(*)::UBIGINT AS reads, sum(ops)::DOUBLE AS ops, sum(len(width))::DOUBLE AS blocks, ",
        "hex(bit_xor(hash(QNAME, FLAG, ref_start, query_start, width))) AS fingerprint ",
        "FROM (SELECT QNAME, FLAG, len(CIGAR) AS ops, %s ",
        "FROM (SELECT QNAME, FLAG, CIGAR, cigar_aligned_blocks(CIGAR, POS) AS b FROM %s))"
      ),
      scalar_blocks, scan_sql(input)
    ),
    stop("unknown workload")
  )
}

same_answer <- function(observed, expected) {
  identical(as.character(observed$reads[[1L]]), as.character(expected$reads[[1L]])) &&
    identical(observed$fingerprint[[1L]], expected$fingerprint[[1L]]) &&
    isTRUE(all.equal(observed$ops[[1L]], expected$ops[[1L]], tolerance = 1e-12)) &&
    isTRUE(all.equal(observed$blocks[[1L]], expected$blocks[[1L]], tolerance = 1e-12))
}

set_affinity <- function(cpus) {
  output <- system2("taskset", c("-pc", cpus, Sys.getpid()), stdout = TRUE, stderr = TRUE)
  if (!is.null(attr(output, "status"))) stop(paste(output, collapse = " "))
  invisible(output)
}

affinity_cpus <- function() {
  line <- system2("taskset", c("-pc", Sys.getpid()), stdout = TRUE)
  spec <- sub("^.*: *", "", line[[length(line)]])
  unlist(lapply(strsplit(spec, ",", fixed = TRUE)[[1L]], function(part) {
    bounds <- as.integer(strsplit(part, "-", fixed = TRUE)[[1L]])
    if (length(bounds) == 1L) bounds else seq.int(bounds[[1L]], bounds[[2L]])
  }))
}
host_cpus <- affinity_cpus()
cpu_one <- as.character(host_cpus[[length(host_cpus)]])
set_affinity(cpu_one)
invisible(dbExecute(con, "SET threads = 1"))

workloads <- c("read_bam only", "blocks in SQL", "cigar_aligned_blocks")
results <- list()
expectations <- list()
for (name in names(inputs)) {
  input <- inputs[[name]]
  expected <- lapply(workloads, function(workload) dbGetQuery(con, workload_sql(input, workload)))
  names(expected) <- workloads
  # Every workload reads the same rows, and the two block derivations must
  # agree on every read's geometry.
  stopifnot(
    length(unique(vapply(expected, function(e) as.character(e$reads[[1L]]), character(1L)))) == 1L,
    identical(expected[["blocks in SQL"]]$fingerprint[[1L]],
              expected[["cigar_aligned_blocks"]]$fingerprint[[1L]]),
    expected[["blocks in SQL"]]$blocks[[1L]] == expected[["cigar_aligned_blocks"]]$blocks[[1L]],
    expected[["cigar_aligned_blocks"]]$blocks[[1L]] > 0
  )
  expectations[[name]] <- expected
  repetitions <- if (name == "ont") 9L else 5L
  for (workload in workloads) {
    sql <- workload_sql(input, workload)
    for (i in seq_len(2L)) stopifnot(same_answer(dbGetQuery(con, sql), expected[[workload]]))
    seconds <- numeric(repetitions)
    for (i in seq_len(repetitions)) {
      started <- proc.time()[["elapsed"]]
      observed <- dbGetQuery(con, sql)
      seconds[[i]] <- proc.time()[["elapsed"]] - started
      stopifnot(same_answer(observed, expected[[workload]]))
    }
    results[[length(results) + 1L]] <- data.frame(
      input = input$label,
      workload = workload,
      threads = 1L,
      affinity = cpu_one,
      repetitions = repetitions,
      reads = as.double(expected[[workload]]$reads[[1L]]),
      ops = expected[[workload]]$ops[[1L]],
      blocks = expected[["cigar_aligned_blocks"]]$blocks[[1L]],
      min_seconds = min(seconds),
      median_seconds = median(seconds),
      max_seconds = max(seconds),
      median_reads_per_second = as.double(expected[[workload]]$reads[[1L]]) / median(seconds),
      fingerprint = expected[[workload]]$fingerprint[[1L]],
      stringsAsFactors = FALSE
    )
  }
}
results <- do.call(rbind, results)
rownames(results) <- NULL

versions <- c(
  duckdb = dbGetQuery(con, "SELECT version() AS v")$v[[1L]],
  htslib = dbGetQuery(con, "SELECT duckhts_htslib_version() AS v")$v[[1L]]
)
invisible(dbDisconnect(con, shutdown = TRUE))

results$revision <- revision
results$run_date <- as.character(Sys.Date())
results$duckdb_version <- versions[["duckdb"]]
results$htslib_version <- versions[["htslib"]]
results$extension_md5 <- unname(tools::md5sum(extension))

dir.create("benchmarks/data", showWarnings = FALSE, recursive = TRUE)
write.csv(results, "benchmarks/data/cigar_aligned_blocks.csv", row.names = FALSE, na = "")

cpu_model <- sub(
  "^model name[[:space:]]*:[[:space:]]*", "",
  grep("^model name", readLines("/proc/cpuinfo"), value = TRUE)[[1L]]
)
ont_receipt <- paste0(inputs$ont$path, ".provenance.tsv")
big <- function(x) format(x, big.mark = ",", scientific = FALSE)
metadata <- data.frame(
  property = c(
    "revision", "src tree", "extension MD5", "run date",
    "long-read input", "long-read reads (ERR14686255)", "long-read ops", "long-read blocks",
    "long-read aligner", "long-read sorter",
    "short-read input", "short-read reads (chr22)", "short-read ops", "short-read blocks",
    "block fingerprint, long-read (SQL = scalar)", "block fingerprint, short-read (SQL = scalar)",
    "DuckDB version", "htslib version", "CPU"
  ),
  value = c(
    revision, src_tree, unname(tools::md5sum(extension)), as.character(Sys.Date()),
    "duckhtsbench registry ont-ecoli-k12: ENA ERR14686255 reads on NCBI RefSeq GCF_000005845.2, minimap2 map-ont, samtools sort",
    big(expectations$ont[["read_bam only"]]$reads[[1L]]),
    big(expectations$ont[["read_bam only"]]$ops[[1L]]),
    big(expectations$ont[["cigar_aligned_blocks"]]$blocks[[1L]]),
    receipt_field(ont_receipt, "aligner"),
    receipt_field(ont_receipt, "sorter"),
    "duckhtsbench registry riker-wgs: 1000 Genomes HG00188 ERR3240174 CRAM transcoded to BAM, region chr22",
    big(expectations$riker[["read_bam only"]]$reads[[1L]]),
    big(expectations$riker[["read_bam only"]]$ops[[1L]]),
    big(expectations$riker[["cigar_aligned_blocks"]]$blocks[[1L]]),
    expectations$ont[["cigar_aligned_blocks"]]$fingerprint[[1L]],
    expectations$riker[["cigar_aligned_blocks"]]$fingerprint[[1L]],
    versions[["duckdb"]], versions[["htslib"]], cpu_model
  ),
  stringsAsFactors = FALSE
)

knitr::kable(metadata, col.names = c("Property", "Recorded value"))
```

## Results

Each workload scans the same rows and returns one row to R, so the timing
covers reading, decoding and the projection rather than transport. Every timed
pass is verified against its own first answer by exact read count, an
order-independent fingerprint, and the summed op and block counts to a relative
tolerance of `1e-12`. The `blocks in SQL` and `cigar_aligned_blocks` workloads
share a fingerprint by construction: it hashes each read's `ref_start`,
`query_start` and `width` lists, and the render asserts that the two
derivations produce the same value on both inputs before timing begins.

```{r results-table, echo=FALSE}
display <- results[, c(
  "input", "workload", "threads", "affinity", "repetitions", "reads", "ops", "blocks",
  "min_seconds", "median_seconds", "max_seconds", "median_reads_per_second"
)]
for (column in c("reads", "ops", "blocks")) {
  display[[column]] <- format(display[[column]], big.mark = ",", scientific = FALSE)
}
for (column in c("min_seconds", "median_seconds", "max_seconds")) {
  display[[column]] <- sprintf("%.3f", display[[column]])
}
display$median_reads_per_second <- format(
  round(display$median_reads_per_second), big.mark = ",", scientific = FALSE
)
names(display) <- c(
  "Input", "Workload", "Threads", "CPU affinity", "Runs", "Reads", "CIGAR ops", "Blocks",
  "Minimum seconds", "Median seconds", "Maximum seconds", "Median reads/s"
)
knitr::kable(display, align = c("l", "l", rep("r", 10)))
```

```{r comparison-table, echo=FALSE}
plain <- results[results$workload == "read_bam only", c("input", "median_seconds")]
names(plain)[2L] <- "plain"
comparison <- merge(
  merge(plain, results[results$workload == "blocks in SQL", c("input", "median_seconds")], by = "input"),
  results[results$workload == "cigar_aligned_blocks", c("input", "median_seconds")],
  by = "input", suffixes = c("_sql", "_scalar")
)
comparison <- comparison[match(unique(results$input), comparison$input), ]
comparison$sql_over_plain <- comparison$median_seconds_sql - comparison$plain
comparison$scalar_over_plain <- comparison$median_seconds_scalar - comparison$plain
comparison$ratio <- comparison$sql_over_plain / comparison$scalar_over_plain
out <- data.frame(
  Input = comparison$input,
  `Plain scan` = sprintf("%.3f", comparison$plain),
  `Blocks in SQL` = sprintf("%.3f", comparison$median_seconds_sql),
  `cigar_aligned_blocks` = sprintf("%.3f", comparison$median_seconds_scalar),
  `SQL cost over plain` = sprintf("%.3f", comparison$sql_over_plain),
  `Scalar cost over plain` = sprintf("%.3f", comparison$scalar_over_plain),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
rownames(out) <- NULL
knitr::kable(out, align = c("l", rep("r", 5)))
```

The cost of each derivation is its median minus the plain scan's median on the
same input, which is the time the projection itself adds. The SQL derivation
grows with the square of each read's op count, so its cost is set by the
long-read input's op distribution; the scalar's cost is linear in ops and is
bounded by the writes of three lists per read.

The measurement does not include staging, extension loading, or any consumer
of the block lists, and it does not measure memory. Both inputs run on one
`read_bam` thread, the long-read input because its genome is one contig and the
short-read input because a region scan is single-threaded, so the numbers say
nothing about scaling across contigs. Small timings vary with scheduling; the
differences are for this host and these inputs only.
